---
title: "R learning from Data and Visual Analytics Lesson 1 Note"
output: html_notebook
---

```{r}
data(iris)
data("faithful")
data(mtcars)
```

```{r}
head(iris)
levels(iris$Species)
```

```{r}
install.packages("ggplot2")
library(ggplot2)
```
The ggplot2 package provides the Fuel Economy (mpg), Midwest County Demographics (midwest) and Diamonds (diamonds) datasets. Once you have installed and loaded ggplot2, you can load the datasets using:
```{r}
data(mpg)
data(midwest)
data(diamonds)
```

```{r}
install.packages("ggplot2movies")
library(ggplot2movies)
data(movies)
```

```{r}
install.packages("plyr")
library(plyr)
data(baseball)
data(ozone)
```

```{r}
library(reshape2) #The reshape2 packages provides some handy functions that you can use to change the structure and format of your data, including the popular melt() function.
data(tips)
```

The MASS package includes a library of advanced statistical modeling functions, and is usually pre-installed with R
```{r}
library(MASS)
```

```{r}
#The robustHD package contains robust statistics and model selection methods for high-dimensional data:
install.packages("robustHD")
```

WRS (optional)
Certain functions in robustHD can be optimized using the WRS package (which depends on WRScpp).

Note: Skip this library if you are getting an error during installation - it is not strictly required for robustHD.
```{r}
# First: Install dependent packages
install.packages(c("MASS", "akima", "robustbase"))

# Second: Install suggested packages
install.packages(c("cobs", "robust", "mgcv", "scatterplot3d", "quantreg", "rrcov", "lars", "pwr", "trimcluster", "parallel", "mc2d", "psych", "Rfit"))

# Third: Install an additional package which provides some C functions
install.packages("devtools")
library(devtools)
install_github("mrxiaohe/WRScpp")  # or install_github("mrxiaohe/WRScppWin") for Windows

# Fourth: Install WRS
install_github("nicebread/WRS", subdir="pkg")
```

```{r}
?tips
head(tips)
summary(tips)
```

